### 2019-03-23 paper/sex_selection_spacing-ver2.tex
Edit regional results for grammar

To understand how regional differences in son preference, fertility, and use of sex 
selection affect birth spacing, I divide India into four broad regions based on degree of 
son preference and initial fertility levels.
Focus is on the most educated women since they are the main users of sex selection.
The four regions are based on \citet{retherford03b}, although expanded to include all 
states in the surveys.
Table \ref{tab:regions} shows the four regions and the states that they contain.%
\footnote{
The geographical names do not perfectly reflect the geography of India, and mainly
serves as mnemonic short-hands.
}
The state names are based on the states in existence when NFHS-1 was collected.
States that are formed later are allocated as closely as possible to the original state.
The ``West'' contains states with a high degree of son preference and use of sex selection.
The ``North'' consists of states that also have a strong son preference, but initially had
higher fertility and, therefore, lower likelihood of using sex selection.
The ``East'' states generally have only a moderate level of son preference.
Finally, the ``South'' contains states with traditionally low son preference and lower
fertility than the rest of India.

@@

Figure \ref{fig:bootstrap_high_r1} shows the predicted average birth intervals, sex 
ratios, and likelihoods of having a child by the end of the spell over time for the three 
spells.
What most stands out is how close to the national estimates the ``West'' estimates are.
The predicted second spell average birth spacing is similar across families with a girl
or a boy, and the average spacing substantially higher in the absence of boys than when
there is at least one boy for the third spell.
The main difference is that the very high predicted sex ratios reflect the expected higher 
use of sex selection.
For the second spell, the predicted sex ratios are around 60 percent for both urban and
rural women, and for the third spell it it around 70 percent for urban women and 
consistently above 60 percent for rural women.
This higher use of sex selection, in turn, translates into longer average spacing.
Furthermore, fertility has fallen faster than the national trend, especially for women
with one or more sons.

@@

Figure \ref{fig:bootstrap_high_r2} shows the same outcomes for the ``North'' region. 
Despite the idea that the ``North'' is characterized by high fertility and lower 
likelihood of sex selection the differences to the ``West'' region is surprisingly small,
at least in the urban areas.
Fertility is, indeed, higher than in the ``West'' region, but the declines in the 
probability of parity progression has been relatively rapid and mirrors the national one.
The use of sex selection is also close to the ``West'', and there is the same reversal of
spacing pattern as there.
The rural areas of the ``North'' are a slightly different story.
Fertility is substantially higher, and the use of sex selection much lower than in the
rural areas of the ``West''. 
That said, there still is clear evidence that the sex ratio is substantially above normal
at close to 60 percent for the latest decade.

@@

The results for the ``East'' region of India are shown in Figure 
\ref{fig:bootstrap_high_r3}, although the fourth spell results are not included because 
the sample size is smaller for this region and the fourth spell results very noisy.
Despite the ``South'' region generally considered to be the low fertility region, the
reduction in fertility has been even faster in the ``East''.
For the last decade only half of urban women with a first-born son are predicted to have
a second child by the end of the spell period covered, which means almost nine years after
the first birth.
The probability is only slightly higher for those in urban areas with a first-born 
daughter and still less than 70 percent.
Not surprisingly, the probabilities are higher in rural areas, but still below the 
national pattern.
Similarly, conditionally on having a second child, the probabilities of having a third
child are low, even in the absence of sons.
In urban areas a woman with two girls have only around a 40 percent likelihood of 
having a third child.

What really stands out, however, is the rapid increase in the lengths of birth intervals,
especially for the second spell.
In both urban and rural areas, the predicted average birth intervals have gone up by
more than a year across the four decade, and that is independent of whether the first-born
was a boy or a girl.
The increases in birth intervals are slightly smaller for third spell, but still 
substantial.
Although spacing pattern is consistently with some son preference with the average
spacing slightly shorter in the absence of boys than when at least one boy is present,
there is little evidence of sex-selective abortions.
The sex ratios after a first-born daughter is slightly elevated, but much lower than 
the prior two regions.
There is more evidence for the third spell, where the sex ratio is close to 60 percent
in the last period for urban women and the last two for rural women, but recall that 
there is a much smaller likelihood of even making it to the third spell for this region
than the other or the national average.

@@

The final region is the ``South'', shown in Figure \ref{fig:bootstrap_high_r4}.
This region is historically thought to have the lowest degree of son preference of the
regions of India.
As for the ``East'', I do not show the fourth spell results because of the small sample 
size and the associated noisy estimates.
Judging from the second spell progression rate and the associated spacing pattern, it
does, indeed, appear that there is little son preference in this region.
The likelihood of having a second birth is relatively high at around 80 to 90 percent,
and is close to independent of the sex of the first birth in both urban and rural areas.
The predicted average spacing also show no sign of son preference, and the predicted
sex ratios are close to the natural rate.
The average birth spacing started high relatively to the other regions, and with little 
change in fertility and no significant use of sex selection, it has increased only 
slightly over time.

With the high likelihood of a second birth and lower decline in the likelihood of a third
birth, the ``South'' has a higher predicted fertility than the ``East''.
What is more, for urban areas the average birth interval have increased substantially
for those with only girls relative to those with a boy and a girl, corresponding to
a predicted sex ratio close to 60 percent with only girls. 
The signs of the use of sex selection also show up for rural women with no sons.
Furthermore, over the four decades the likelihood of having a third birth has decreased
much less than for either a boy and a girl or two boys.%
\footnote{
The relatively high parity progression for women without son does raise the question
whether ``daughter avoidance'' is behind the fertility behavior in this region as
suggested by \citet{DiamondSmith2008}.
}
The differences in parity progression and the sex ratios suggests that despite originally
having less son preference, the ``South'' now appears to have the longer average spacing 
in absence of son and the higher sex ratio associated with the use of sex selection.

>>>>

To understand how regional differences in son preference, fertility, and use of sex 
selection affect birth spacing, I divide India into four broad regions based on the degree 
of son preference and initial fertility levels.
The focus is on the most educated women since they are the primary users of sex selection.
The four regions are based on \citet{retherford03b}, although expanded to include all 
states in the surveys.
Table \ref{tab:regions} shows the four regions and the states that they contain.%
\footnote{
The geographical names do not entirely reflect the geography of India, and mainly
serves as mnemonic short-hands.
}
The state names are those in existence at the time of the NFHS-1 data collection.
States formed later are allocated as close as possible to the original state.
The ``West'' contains states with a high degree of son preference and use of sex selection.
The ``North'' consists of states that also have a strong son preference, but initially had
higher fertility and, therefore, lower likelihood of using sex selection.
The ``East'' states generally have only a moderate level of son preference.
Finally, the ``South'' contains states with traditionally low son preference and lower
fertility than the rest of India.

@@

Figure \ref{fig:bootstrap_high_r1} shows the predicted average birth intervals, sex 
ratios, and likelihoods of having a child by the end of the spell over time for the three 
spells in the ``West.''
What most stands out is how close to the national estimates the ``West'' estimates are.
The predicted second spell average birth spacing is similar across families with a girl
or a boy, and the average spacing substantially higher in the absence of boys than when
there is at least one boy for the third spell.
The main difference is that the very high predicted sex ratios reflect the expected higher 
use of sex selection.
For the second spell, the predicted sex ratios are around 60 percent for both urban and
rural women, and for the third spell it is around 70 percent for urban women and 
consistently above 60 percent for rural women.
This higher use of sex selection, in turn, translates into longer average spacing.
Furthermore, fertility has fallen faster than the national trend, especially for women
with one or more sons.

@@

Figure \ref{fig:bootstrap_high_r2} shows the same outcomes for the ``North'' region. 
Despite the idea that the ``North'' has high fertility and lower 
likelihood of sex selection the differences to the ``West'' region is surprisingly small,
at least in the urban areas.
Fertility is, indeed, higher than in the ``West'' region, but the declines in the 
probability of parity progression have been relatively rapid and mirror the national one.
The use of sex selection is also close to the ``West,'' and there is the same reversal of
spacing pattern as there.
The rural areas of the ``North'' are a slightly different story.
Fertility is substantially higher, and the use of sex selection much lower than in the
rural areas of the ``West.'' 
That said, there still is clear evidence that the sex ratio is substantially above normal
at close to 60 percent for the latest decade.

@@

The results for the ``East'' region of India are shown in Figure 
\ref{fig:bootstrap_high_r3}, although it does not include the fourth spell results because 
the sample size is smaller for this region and the fourth spell results very noisy.
Despite the ``South'' region generally considered to be the low fertility region, the
reduction in fertility has been even faster in the ``East.''
For the last decade, only half of the urban women with a first-born son are predicted to 
have a second child by the end of the spell, which means almost nine years after the first 
birth.
The probability is only slightly higher for those in urban areas with a first-born 
daughter and still less than 70 percent.
Not surprisingly, the probabilities are higher in rural areas, but still below the 
national pattern.
Similarly, conditionally on having a second child, the probabilities of having a third
child are low, even in the absence of sons.
In urban areas, a woman with two girls has only around a 40 percent likelihood of 
having a third child.

What stands out, however, is the rapid increase in the lengths of birth intervals,
especially for the second spell.
In both urban and rural areas, the predicted average birth intervals have gone up by
more than a year across the four decades, and that is independent of whether the first-born
was a boy or a girl.
The increases in birth intervals are slightly smaller for the third spell but still 
substantial.
Although the spacing pattern is consistent with some son preference, with the average
spacing slightly shorter in the absence of boys than when at least one boy is present,
there is little evidence of sex-selective abortions.
The sex ratios after a first-born daughter are slightly elevated but much lower than 
the prior two regions.
There is more evidence for the third spell, where the sex ratio is close to 60 percent
in the last period for urban women and the last two for rural women, but recall that 
there is a much smaller likelihood of even making it to the third spell for this region
than the other or the national average.

@@

The final region is the ``South,'' shown in Figure \ref{fig:bootstrap_high_r4}.
This region is historically thought to have the lowest degree of son preference of the
regions of India.
As for the ``East,'' I do not show the fourth spell results because of the small sample 
size and the associated noisy estimates.
Judging from the second spell progression rate and the associated spacing pattern, it
does, indeed, appear that there is little son preference in this region.
The likelihood of having a second birth is relatively high at around 80 to 90 percent
and is close to independent of the sex of the first birth in both urban and rural areas.
The predicted average spacing also show no sign of son preference, and the predicted
sex ratios are close to the natural rate.
The average birth spacing started high relative to the other regions, and with little 
change in fertility and no significant use of sex selection, it has increased only 
slightly over time.

With the high likelihood of a second birth and lower decline in the likelihood of a third
birth, the ``South'' has higher predicted fertility than the ``East.''
What is more, for urban areas the average birth interval has increased substantially
for those with only girls relative to those with a boy and a girl, corresponding to
a predicted sex ratio close to 60 percent with only girls. 
The signs of the use of sex selection also show up for rural women with no sons.
Furthermore, over the four decades, the likelihood of having a third birth has decreased
much less than for either a boy and a girl or two boys.%
\footnote{
The relatively high parity progression for women without son does raise the question of
whether ``daughter avoidance'' is behind the fertility behavior in this region as
suggested by \citet{DiamondSmith2008}.
}
The differences in parity progression and the sex ratios suggest that despite originally
having less son preference, the ``South'' now appears to have longer average spacing 
in the absence of sons and the higher sex ratio associated with the use of sex selection.
